---
title: "AI Enhanced CJ"
subtitle: "A Natural Experiment"
author: "Chris Wheadon"
institute: "No More Marking"
format:
  revealjs:
    theme: serif
    logo: Master-NMM.png
    fontsize: 28px
---

```{=html}
<style>
.reveal h2 {
  background-color: #3C1360;
  color: #ffffff;
  padding: 0.2em 0.4em;
  margin-bottom: 0.5em;
  border-radius: 4px;
}
.reveal h3 {
  color: #00C1B4;
  border-bottom: 2px solid #00C1B4;
  padding-bottom: 0.1em;
}
.reveal blockquote {
  border-left: 4px solid #F26522;
  color: #2D2D2D;
}
.reveal .callout {
  border-left-color: #3C1360;
}
.reveal .no-header h2 {
  display: none;
}
.reveal h2:empty {
  display: none;
}
</style>
```

```{r}
#| warning: false
#| echo: false
#| include: false

rm(list=ls())
require(tidyverse)
require(here)
require(knitr)
require(rstatix)
require(ggpubr)

final_data <- read_csv(here("output", "apw_feb_to_oct_correlation_data_anon.csv"))
final_data <- final_data %>%
  mutate(ai_user = factor(ai_user, levels = c(FALSE, TRUE), labels = c("Non AI User", "AI User")))
final_data <- final_data %>%
  rename_with(~ str_replace(., "_base", "_feb_year_2")) %>%
  rename_with(~ str_replace(., "_target", "_oct_year_3"))

mean_feb <- mean(final_data$finalScore_feb_year_2)
mean_oct <- mean(final_data$finalScore_oct_year_3)
lm_model <- lm(finalScore_oct_year_3 ~ finalScore_feb_year_2, data = final_data)
final_data <- final_data %>%
  mutate(predicted_oct = predict(lm_model),
         progress = finalScore_oct_year_3 - predicted_oct,
         progress_level = case_when(
           finalScore_oct_year_3 >= mean_oct & progress >= 0 ~ "Above Mean, More Progress",
           finalScore_oct_year_3 >= mean_oct & progress < 0 ~ "Above Mean, Less Progress",
           finalScore_oct_year_3 < mean_oct & progress >= 0 ~ "Below Mean, More Progress",
           finalScore_oct_year_3 < mean_oct & progress < 0 ~ "Below Mean, Less Progress"
         ))
```

## Context

- APW 3 is the writing of pupils in Year 3, aged 7 to 8
- Use this picture to write a story
- Handwritten
- In Feb of year 2 no AI was available, teachers had to judge
- In Oct of year 3, schools could choose whether or not to use AI judges
- The following analysis considers **matched pupils** between the two sessions

---

## Measuring Progress

- We use an anchoring design where scripts are repeated between sessions
- We calculate a shift constant between sessions based on the shift in scores from the anchors
- Same anchors for AI and non-AI route

---

## Descriptive Statistics

The February results of the two groups were very similar. After using AI in October, the mean scores were again similar — but the **spread narrowed for AI users**.

```{r}
#| echo: false
#| label: tbl-descriptives
#| tbl-cap: "Descriptive Statistics by AI User Status"

descriptive_stats <- final_data %>%
  group_by(ai_user) %>%
  summarise(
    mean_feb_year_2 = mean(finalScore_feb_year_2),
    sd_feb_year_2 = sd(finalScore_feb_year_2),
    mean_oct_year_3 = mean(finalScore_oct_year_3),
    sd_oct_year_3 = sd(finalScore_oct_year_3),
    n = n()
  )
descriptive_stats %>%
  knitr::kable(
    digits = 0,
    col.names = c("AI User", "Mean", "SD", "Mean", "SD", "N")
  ) %>%
  kableExtra::add_header_above(c(" " = 1, "Feb Year 2" = 2, "Oct Year 3" = 2, " " = 1))
```

---

## Score Distributions

```{r}
#| echo: false
#| warning: false
#| label: fig-score-distributions
#| fig-cap: "Distributions of Final Scores by AI User Status"

final_data %>%
  pivot_longer(cols = c(finalScore_feb_year_2, finalScore_oct_year_3),
               names_to = "assessment", values_to = "score") %>%
  ggplot(aes(x = score, fill = ai_user)) +
  geom_density(alpha = 0.5) +
  xlim(300, 700) +
  labs(x = "Final Score", y = "Density", fill = "AI User in Oct") +
  theme_minimal() +
  facet_wrap(~ assessment,
             labeller = as_labeller(c(`finalScore_feb_year_2` = "APW Feb",
                                      `finalScore_oct_year_3` = "APW Oct")))
```

---

## Statistical Tests

```{r}
#| echo: false
#| warning: false
#| label: tbl-statistical-tests

long_data <- final_data %>% select(
  student,
  finalScore_feb_year_2,
  finalScore_oct_year_3,
  ai_user 
) %>% pivot_longer(
  cols = c(finalScore_feb_year_2, finalScore_oct_year_3),
               names_to = "assessment", values_to = "score"
)

res.anova <- long_data %>%
  anova_test(dv = score, wid = student, between = ai_user, within = assessment)
get_anova_table(res.anova) |>
  knitr::kable(digits = 3)
```

---

## Statistical Tests: Interaction

```{r}
#| echo: false
#| warning: false
#| label: fig-statistical-tests

# Create the interaction plot
ggline(long_data, 
       x = "assessment", 
       y = "score", 
       color = "ai_user", 
       add = c("mean_sd"),
       palette = "jco",
       ylab = "Test Score", 
       xlab = "Testing Period",
       title = "Impact of Condition on Student Performance")
```

---

## Correlations

Regression lines show average progress from February to October was **similar regardless of AI use**.

```{r}
#| echo: false
#| warning: false
#| label: fig-score-correlation

final_data %>%
  ggplot(aes(x = finalScore_feb_year_2, y = finalScore_oct_year_3)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ggpmisc::stat_poly_eq(
    aes(label = paste(after_stat(eq.label), after_stat(rr.label), sep = "~~~")),
    formula = y ~ x, parse = TRUE, size = 3
  ) +
  ylim(300, 700) +
  labs(title = "Correlation of Final Scores from APW Feb to APW Oct",
       x = "Final Score in APW Feb",
       y = "Final Score in APW Oct") +
  theme_minimal() +
  facet_wrap(~ ai_user)
```

---

## Correlations by Group

Correlations for AI users were **higher** than for non-AI users, though both were strong.

```{r}
#| echo: false
#| warning: false
#| label: tbl-correlations
#| tbl-cap: "Correlation between APW Feb and APW Oct Final Scores by AI User Status"

correlation_results <- final_data %>%
  group_by(ai_user) %>%
  summarise(n = n(), correlation = cor(finalScore_feb_year_2, finalScore_oct_year_3)) %>%
  ungroup()
knitr::kable(correlation_results, digits = 2)
```

---

## Regression to the Mean & Progress Levels

AI users had proportionally more pupils with **above-average scores making less than average progress** — a more pronounced regression to the mean effect.

```{r}
#| echo: false
#| warning: false
#| label: fig-progress-levels
#| fig-cap: "Progress Levels from APW Feb to APW Oct by AI User Status"

final_data %>%
  ggplot(aes(x = finalScore_feb_year_2, y = finalScore_oct_year_3, color = progress_level)) +
  geom_point(alpha = 0.5) +
  ylim(300, 700) +
  labs(x = "Final Score in APW Feb",
       y = "Final Score in APW Oct",
       color = "Progress Level") +
  theme_minimal() +
  facet_wrap(~ ai_user)
```

---

## Progress Levels: Summary

```{r}
#| echo: false
#| warning: false
#| label: tbl-progress-levels
#| tbl-cap: "Progress Levels from APW Feb to APW Oct by AI User Status (% of pupils)"

progress_summary <- final_data %>%
  group_by(ai_user, progress_level) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(ai_user) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

progress_summary %>%
  pivot_wider(names_from = progress_level, values_from = c(n, pct), values_fill = 0) %>%
  select(ai_user, starts_with("pct")) %>%
  rename_with(~ str_remove(., "^pct_")) |>
  (\(df) knitr::kable(df, digits = 2, col.names = c("", names(df)[-1])))()
```

---

## Number of Comparisons

- AI users completed **substantially more comparisons** → likely more reliable measurement

```{r}
#| echo: false
#| warning: false
#| label: tbl-comparisons-summary
#| tbl-cap: "Mean Comparisons Per Script by AI User Status"

infit_summary <- final_data %>%
  group_by(ai_user) %>%
  summarise(
    mean_comparisons_feb_year_2 = mean(comparisons_feb_year_2, na.rm = TRUE),
    mean_comparisons_oct_year_3 = mean(comparisons_oct_year_3, na.rm = TRUE),
    pct_feb_year_2 = sum(infit_feb_year_2 > 1.3, na.rm = TRUE) / n() * 100,
    pct_oct_year_3 = sum(infit_oct_year_3 > 1.3, na.rm = TRUE) / n() * 100
  )

knitr::kable(
  infit_summary %>% select(ai_user, mean_comparisons_feb_year_2, mean_comparisons_oct_year_3),
  digits = 2,
  col.names = c("", "Feb Year 2", "Oct Year 3")
)
```

---

## Infit Summary

- AI users had a **much lower % of pupils with infit > 1.3**
- Note: infit is sensitive to number of comparisons, so this doesn't necessarily reflect more consistent judging

```{r}
#| echo: false
#| warning: false
#| label: tbl-infit-summary
#| tbl-cap: "% of Pupils with Infit > 1.3 by AI User Status"

knitr::kable(
  infit_summary %>% select(ai_user, pct_feb_year_2, pct_oct_year_3),
  digits = 2,
  col.names = c("", "Feb Year 2", "Oct Year 3")
)
```

---

## Issues: Hallucinated Transcriptions

- The AI occasionally produces transcriptions that do not match the original handwritten script
- The **Fall of the House of Usher** issue
- Handwriting degrades rapidly under exam conditions
- Requires mutliple models & post-hoc checking to identify and remove affected scripts

---

## Issues: Weighting in Training Data

- The AI's training data may over-represent certain writing styles or authors
- The **Carol Ann Duffy issue**: if you are writing about a poem by Carol ...
- This could introduce systematic bias in scores that is not present in human judging

---

## Issues: Unexpected Responses

- In some cases the AI produces responses outside the expected scoring range or format
- The **One Day issue**: "1 Day."
- Very short responses when an essay is expected cause unexpected results

---

## Conclusion

::: columns
::: {.column style="width: 50%; font-size: 0.85em;"}
### What we found
- Similar average progress for both groups
- Strong correlations in both groups (higher for AI users)
- Regression to the mean more pronounced for AI users
- More comparisons completed by AI → likely more reliable measurement
:::
::: {.column style="width: 50%; font-size: 0.85em;"}
### What remains open
- Does AI reward different skills or writing approaches than human judges?
- Further qualitative analysis needed to explore this question
:::
:::
